---
title: "R Data Cleaning Tutorial - Breaking the Data"
output:
  html_document:
    df_print: paged
---

This notebook is designed to provide a replicable process for building a dataset that contains issues that will then be used in the demonstration of methods for "cleaning" data prior to analysis 

The data for this demonstration are based upon the [idigbio_rodents.csv](https://figshare.com/articles/idigbio_rodents_csv/5535724) dataset. The data are described as follows in the repository where they are shared:

> The idigbio_rodents.csv dataset is just over 10k records and contains data from natural history collections specimen records representing 5 genera from 4 major US collections, limited to US records. All records are from the Order Rodentia. All the data are mapped to the biodiversity data standard Darwin Core (http://rs.tdwg.org/dwc/terms/).

The requested citation for these data is:

Collins, Matthew; Paul, Deborah (2017): idigbio_rodents.csv. figshare. Dataset. [https://doi.org/10.6084/m9.figshare.5535724.v1](https://doi.org/10.6084/m9.figshare.5535724.v1) 


Import the source data and save a local copy
```{r}
rawData <- read.csv("https://ndownloader.figshare.com/files/9582724")
write.csv(rawData, file = "../data/rawData.csv")
```

Generate some DMS (degree-minute-second) versions of the provided DD (decimal degrees) columns
```{r}
if (exists("learning")) {rm("learning")}
learning <- rawData
# calculate the degree portion
learning$latDeg <- unlist(floor(learning['decimalLatitude']))
# calculate the minute portion
learning$latMin <- floor((learning['decimalLatitude'] - learning$latDeg) * 60)
# calculate the second portion
learning$latSec <- round((((learning['decimalLatitude'] - learning$latDeg) - (learning$latMin/60)) * 3600), 3)
# jam them together into a lovely but not very useful text string
learning$latDMS <- apply(learning, 1, function(x) paste(x['latDeg'],"°", x['latMin'],"'", x['latSec'], "\"N", sep=""))

# do it agree for the longitude values
learning$lonDeg <- floor(learning['decimalLongitude'])
learning$lonMin <- floor((learning['decimalLongitude'] - learning$lonDeg) * 60)
learning$lonSec <- round((((learning['decimalLongitude'] - learning$lonDeg) - (learning$lonMin/60)) * 3600), 3)
learning$lonDMS <- apply(learning, 1, function(x) paste(x['lonDeg'],"°", x['lonMin'],"'", x['lonSec'], "\"E", sep=""))

# get rid of the intermediate deg, min, sec columns used to create the composite DMS columns
learning <- subset(learning, select = -c(latDeg, latMin, latSec, lonDeg, lonMin, lonSec))
```

Generate a column of mixed data types
```{r}
# convert the DD lat values to character strings representing the numbers
learning$textLatDD <- as.character(learning$decimalLatitude)
# replace lovely NA values with "missing" strings
learning$textLatDD[is.na(learning$textLatDD)] <- "missing"
```

Write out the learning data file
```{r}
write.csv(learning, file = "../data/learning.csv")
```