---
title: "R Data Cleaning Tutorial - Breaking the Data"
output: html_notebook
---

This notebook is designed to provide a replicable process for building a dataset that contains issues that will then be used in the demonstration of methods for "cleaning" data prior to analysis 


Import the source data and save a local copy
```{r}
rawData <- read.csv("https://ndownloader.figshare.com/files/9582724")
write.csv(rawData, file = "../data/rawData.csv")
```

Generate some DMS versions of the provided DD columns
```{r}
rm("learning")
learning <- rawData
learning$latDeg <- unlist(floor(learning['decimalLatitude']))
learning$latMin <- floor((learning['decimalLatitude'] - learning$latDeg) * 60)
learning$latSec <- round((((learning['decimalLatitude'] - learning$latDeg) - (learning$latMin/60)) * 3600), 3)
learning$latDMS <- apply(learning, 1, function(x) paste(x['latDeg'],"°", x['latMin'],"'", x['latSec'], "\"N", sep=""))

learning$lonDeg <- floor(learning['decimalLongitude'])
learning$lonMin <- floor((learning['decimalLongitude'] - learning$lonDeg) * 60)
learning$lonSec <- round((((learning['decimalLongitude'] - learning$lonDeg) - (learning$lonMin/60)) * 3600), 3)
learning$lonDMS <- apply(learning, 1, function(x) paste(x['lonDeg'],"°", x['lonMin'],"'", x['lonSec'], "\"E", sep=""))
learning <- subset(learning, select = -c(latDeg, latMin, latSec, lonDeg, lonMin, lonSec))
```

Generate a column of mixed data types
```{r}
#learning <- subset(learning, select = -c(textLatDD))
learning$textLatDD <- as.character(learning$decimalLatitude)
learning$textLatDD[is.na(learning$textLatDD)] <- "missing"
```

Write out the learning data file
```{r}
write.csv(learning, file = "../data/learning.csv")
```